The evolution of the transcriptome of mammalian placentas

### 2016-09-20 manuscript/manuscript.Rnw
start discussing shape

In order to address our two questions in a more systematic way, we
sequenced and quantified the mRNA environment of mammalian placentas
at term, including that of the \human, \textit{Pan paniscus} (the
sister species to \textit{P. troglodytes}), \textit{Ateles fusciceps},
\mouse, \galili, \carmeli, \textit{Canis familiaris}, \cow, \elephant,
\textit{Dasypus novemcinctus}, and \textit{Monodelphis domesticus}.
These species represent four major branches of eutherian mammals
(Euarchontoglires, Laurasiatheria, Afrotheria, and Xenartha) as well
as marsupials. These placentas also represent a wide array of
placental morphologies as well as traits such as number and size of
offspring and length of gestation (\cref{tab:gestational_periods}). We
use this data, along with existing RNA-Seq data from \mouse
\cite{Wang.ea2011:surveyfornovelimprinted}, \human
\cite{Hughes.ea2015:Evaluatingintra-andinter-individual},
\textit{Equus caballus}
\cite{Coleman.ea2010:Structuralannotationofequine}, \textit{Ovis
  aries} \cite{Archibald.ea2010:sheepgenomereferencesequence}, and
\textit{Sus scrofa} \cite{Ahanda.ea2012:Predictionofaltered3-} to: 1)
identify the non-housekeeping genes which are expressed in the
placentas of all species analyzed and therefore represent the
\emph{core functionality of the placenta}; 2) identify the genes whose
expression level changes significantly in each of the morphology
types.

@@

The
\Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
genes comprising the core placenta transcriptome are shown in
\cref{fig:core_placenta_transcriptome} ordered by their median
transcription level in \ac{FPKM}. We considered a gene to be core to
the placenta transcriptome if a 1:1 ortholog of that gene was
expressed with $\text{FPKM} \ge 10$ in all species studied and was
not a human housekeeping gene according to
\citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}.
Supplemental \cref{fig:core_placenta_transcriptome_shape} shows the number of genes
which would be included in the core placenta set if the \ac{FPKM}
cutoff was set differently.
%
Multiple components of annexin complexes are included in the core
transcriptome, including \textit{ANXA2}, \textit{ANXA1},
\textit{S100A11}, \textit{S100A10}. ANXA1 is involved in the
resolution of the inflammation response through inhibition of
phospholipase A2 activity and signaling through the formyl peptide
receptor family, and has previously been hypothesized to be important
for the maintenance of the anti-inflammatory state during pregnancy
\cite{Hutchinson.ea2011:Molecularregulatorsofresolution}. ANXA2 is
known to interact with S100A10, S100A6, S100A11, and S100A4
\cite{Liu.ea2015:AnnexinA2complexeswith}, and there is evidence for
its interaction with S100P as well; it is involved in cell-cell
interaction as well as vesicle trafficking and von Willibrand Factor
secretion. The expression regulation of another core gene,
\textit{TIMP3}, has been implicated in pre-eclampsia
\cite{Ma.ea2014:Down-regulationofTIMP3leads,Xiang.ea2013:PromoterhypomethylationofTIMP3}.
Supplemental \cref{tab:core_placenta_go_terms} lists the significant
enrichment of \ac{GO} terms in the core placenta transcriptome;
notable terms include those involved in cytoskeletal structure,
protein transport, membrane protein structure, and inflammation
resolution. Increasing the permissiveness of the core placenta
transcriptome by only requiring 10 of 14 species to express 1:1
orthologs with FPKM $>10$ and results in the same terms also being
significantly enriched, along with additional terms (Supplemental
\cref{tab:core_placenta_go_terms}).

@@

\subsection{\ac{GSE} Analyses of Individual Transcriptomes}

In order to identify major functional categories of genes expressed in
the individual transcriptomes, we examined \ac{GSE} in 1:1 human
orthologs which had \ac{FPKM} values in the top 1\% of expressed genes
(Supplemental %
\cref{tab:go_terms_enriched_top_one_dasypus_novemcinctus,tab:go_terms_enriched_top_one_equus_caballus,tab:go_terms_enriched_top_one_homo_sapiens,tab:go_terms_enriched_top_one_loxodonta_africana,tab:go_terms_enriched_top_one_monodelphis_domestica,tab:go_terms_enriched_top_one_mus_musculus,tab:go_terms_enriched_top_one_ovis_aries,tab:go_terms_enriched_top_one_pan_paniscus,tab:go_terms_enriched_top_one_sus_scrofa,tab:go_terms_enriched_top_one_ateles_fusciceps,tab:go_terms_enriched_top_one_spalax_galili,tab:go_terms_enriched_top_one_spalax_carmeli,tab:go_terms_enriched_top_one_bos_taurus,tab:go_terms_enriched_top_one_canis_familiaris}) using topGO \cite{topGO:2.20.0}. %
%
Significant enrichment in genes ($\text{FDR}\le0.05$) in the top 1\%
of \ac{FPKM} values in at least 12 of the
\Sexpr{length(species.names)} species included clusters such as
membrane organization , locomotion, tissue development, extracellular
matrix organization, regulation of peptidase activity, and various
cell death related pathways among others
(Supplemental \cref{tab:go_terms_all_species}). If we instead examine the top 100
genes by \ac{FPKM} in each of the \Sexpr{length(species.names)}
species (Supplemental %
\cref{tab:go_terms_enriched_dasypus_novemcinctus,tab:go_terms_enriched_equus_caballus,tab:go_terms_enriched_homo_sapiens,tab:go_terms_enriched_loxodonta_africana,tab:go_terms_enriched_monodelphis_domestica,tab:go_terms_enriched_mus_musculus,tab:go_terms_enriched_ovis_aries,tab:go_terms_enriched_pan_paniscus,tab:go_terms_enriched_sus_scrofa,tab:go_terms_enriched_ateles_fusciceps,tab:go_terms_enriched_spalax_galili,tab:go_terms_enriched_spalax_carmeli,tab:go_terms_enriched_bos_taurus,tab:go_terms_enriched_canis_familiaris})
notable enrichments include extracellular matrix, female pregnancy,
integrin and collagen-related processes, and various cell death
related processes.

>>>>

In order to address our two questions in a more systematic way, we
sequenced and quantified the total RNA of mammalian placentas at term,
including that of the \human, \textit{Pan paniscus} (the sister
species to \textit{P. troglodytes}), \textit{Ateles fusciceps},
\mouse, \galili, \carmeli, \textit{Canis familiaris}, \cow, \elephant,
\textit{Dasypus novemcinctus}, and \textit{Monodelphis domesticus}.
These species represent four major branches of eutherian mammals
(Euarchontoglires, Laurasiatheria, Afrotheria, and Xenartha) as well
as marsupials. These placentas also represent a wide array of
placental morphologies as well as traits such as number and size of
offspring and length of gestation (\cref{tab:gestational_periods}). We
use this data, along with existing RNA-Seq data from \mouse
\cite{Wang.ea2011:surveyfornovelimprinted}, \human
\cite{Hughes.ea2015:Evaluatingintra-andinter-individual},
\textit{Equus caballus}
\cite{Coleman.ea2010:Structuralannotationofequine}, \textit{Ovis
  aries} \cite{Archibald.ea2010:sheepgenomereferencesequence}, and
\textit{Sus scrofa} \cite{Ahanda.ea2012:Predictionofaltered3-} to: 1)
identify the non-housekeeping genes which are expressed in the
placentas of all species analyzed and therefore represent the
\emph{core functionality of the placenta}; 2) identify the genes whose
expression level changes significantly in each of the morphology
types.

@@

The
\Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
genes comprising the core placenta transcriptome are shown in
\cref{fig:core_placenta_transcriptome} ordered by their median
transcription level in \ac{FPKM}. We considered a gene to be core to
the placenta transcriptome if a 1:1 ortholog of that gene was
expressed with $\text{FPKM} \ge 10$ in all species studied and was
not a human housekeeping gene according to
\citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}.
\cref{fig:core_placenta_transcriptome_shape} shows the number of genes
which would be included in the core placenta set if the \ac{FPKM}
cutoff was set differently; for example, choosing an \ac{FPKM} cutoff of 1 yields %
\Sexpr{cpt.distribution.shape[`Expression (FPKM)`>1 & 
`Minimum Expression in` == '14 of 14 species',max(num_genes)]} %
genes in the core transcriptome.
%
Multiple components of annexin complexes are included in the core
transcriptome, including \textit{ANXA2}, \textit{ANXA1},
\textit{S100A11}, \textit{S100A10}. ANXA1 is involved in the
resolution of the inflammation response through inhibition of
phospholipase A2 activity and signaling through the formyl peptide
receptor family, and has previously been hypothesized to be important
for the maintenance of the anti-inflammatory state during pregnancy
\cite{Hutchinson.ea2011:Molecularregulatorsofresolution}. ANXA2 is
known to interact with S100A10, S100A6, S100A11, and S100A4
\cite{Liu.ea2015:AnnexinA2complexeswith}, and there is evidence for
its interaction with S100P as well; it is involved in cell-cell
interaction as well as vesicle trafficking and von Willibrand Factor
secretion. The expression regulation of another core gene,
\textit{TIMP3}, has been implicated in pre-eclampsia
\cite{Ma.ea2014:Down-regulationofTIMP3leads,Xiang.ea2013:PromoterhypomethylationofTIMP3}.
Supplemental \cref{tab:core_placenta_go_terms} lists the significant
enrichment of \ac{GO} terms in the core placenta transcriptome;
notable terms include those involved in cytoskeletal structure,
protein transport, membrane protein structure, and inflammation
resolution. Increasing the permissiveness of the core placenta
transcriptome by only requiring 10 of 14 species to express 1:1
orthologs with FPKM $>10$ and results in the same terms also being
significantly enriched, along with additional terms (Supplemental
\cref{tab:core_placenta_go_terms}).

@@

\subsection{Placenta Morphology and Genetics}

\subsubsection{Placenta Attachment Shape}

\subsubsection{Degree of Intimacy}

\subsubsection{Interdigitation}
